---
title: 'Data exploration and visualisation'
author: "Bernard Silenou and Henrik Schanze"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
---


- After imparting and cleaning your data, the next aim before modelling is to understand your data.
- Essential tools to attain this aim are *summary statistics* and *plots*.

## Goals
- Be able to calculate a range of summary statistics on numeric data
- Be able to draw plots and export out of R

# Summary statistics
- Summary statistics are used to summarize a data in order to communicate the largest amount of information as simply as possible.

- Measures often used are arithmetic mean, variance or standard deviation, and  correlation coefficient.

- The *obama_vs_mccain* dataset contains the fractions of people voting for Obama and McCain in the 2008 US presidential elections.

- We load this data from the R package *learningr* and use it in subsequent examples:

```{r}
data(obama_vs_mccain, package = "learningr")
votesObama <- obama_vs_mccain$Obama
```


## List of summary statistics
Let {x} be the votes for Obama.

Measure of location:

`mean(x)`: sum of x divided by the length
```{r}
mean(votesObama)
```

`median(x)`: 50% of x is above and 50% is below
```{r}
median(votesObama)
```

---



- Measure of variation:
`sd(x)`: standard deviation
`IQR(x)`: interquartile range (robust equivalent of sd when outliers are present in the data)
`mad(x)`: median absolute deviation (robust equivalent of sd when outliers are present in the data)

- Measure of rank:
`min(x)`: minimum value of x
`max(x)`: maximum value of x
`quantile(x, 0.25)`: 25% of x is below this value

- Measure of position:
`first(x)`: equivalent to x[1]
`nth(x, 2)`: equivalent to n<-2; x[n]
`last(x)`: equivalent to x[length(x)]

- Counts:
`n(x)`: the number of element in x
`sum(!is.na(x))`: count non-missing values
`n_distinct(x)`: count the number of unique value






# plots
## Systems of plots

## Scatter plot

## Line plot

## Histogram 

## Bar chart

