---
title: 'Data exploration and visualisation'
author: "Bernard Silenou and Henrik Schanze"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
editor_options: 
  markdown: 
    wrap: 72
---

-   After imparting and cleaning your data, the next aim before
    modelling is to understand your data.
-   Essential tools to attain this aim are *summary statistics* and
    *plots*.

## Goals

-   Be able to calculate a range of summary statistics on numeric data
-   Be able to draw plots and export out of R

# Summary statistics

-   Summary statistics are used to summarize a data in order to
    communicate the largest amount of information as simply as possible.

-   Measures often used are arithmetic mean, variance or standard
    deviation, and correlation coefficient.

-   The *obama_vs_mccain* dataset contains the fractions of people
    voting for Obama and McCain in the 2008 US presidential elections.

-   We load this data from the R package *learningr* and use it in
    subsequent examples:

```{r}
data(obama_vs_mccain, package = "learningr")
votesObama <- obama_vs_mccain$Obama
```

\% \## List of summary statistics

Let {x} be the votes for Obama.

### Measure of location:

`mean(x)`: sum of x divided by the length

```{r}
mean(votesObama)
```

`median(x)`: 50% of x is above and 50% is below

```{r}
median(votesObama)
```

### Measure of rank:

`min(x)`: minimum value of x

```{r}
min(votesObama)
```

`max(x)`: maximum value of x

```{r}
max(votesObama)
```

`quantile(x, 0.25)`: 25% of x is below this value; `quantile(x, 0.50)`==
median(x)

```{r}
quantile(votesObama, 0.50)
median(votesObama)
quantile(votesObama)
```

### Measure of variation:

`sd(x)`: standard deviation, square sd to get variance or use `var(x)`

```{r}
sd(votesObama)
var(votesObama) 
sd(votesObama)**2
```

`range(x)`: gives min and max values

```{r}
range(votesObama)

```

`IQR(x)`: interquartile range == 75th percentile minus 25th percentile

```{r}
IQR(votesObama)

```

### Measure of position or count:

`x[1]`: first element in x

```{r}
votesObama[1]
```

`x[length(x)]`: last element in x

```{r}
votesObama[length(votesObama)]
```

**Quick exercise**: what is the last last but one element in
`votsObama`?

### Other summary statistics functions:

`summary(x)`: generic function to compute multiple statistics at once

```{r}
summary(votesObama)
summary(obama_vs_mccain)
```

`length(x)`: the number of element in x

```{r}
length(votesObama)
dim(obama_vs_mccain)
dim(obama_vs_mccain)[1] # the same as nrow(obama_vs_mccain)
dim(obama_vs_mccain)[2] # the same as ncol(obama_vs_mccain)
```

`table(is.na(x))`: count the number of missing and non missing values

```{r}
table(is.na(votesObama))
```

`length(unique(x))`: count the number of unique value

```{r}
length(unique(obama_vs_mccain$Region))
```

`cor(x)`: calculates correlations between numeric vectors, "pearson"
(default)

```{r}
cor(obama_vs_mccain$Obama, obama_vs_mccain$McCain)
```

`cov(x)`: calculates covariance between numeric vectors, "pearson"
(default)

```{r}
with(obama_vs_mccain, cov(Obama, McCain)) # the same as cov(obama_vs_mccain$Obama, obama_vs_mccain$McCain)
```

# Plots

## Systems of plots

R has three systems of graphics: *base*, *grid*, and *lattice*.

**Base graphics** is easy to get started with but hard to extend to new
graph types. It will be used for a greater part of this section.

**Grid graphics** provides greater flexibility but require writing much
lines of code. Example is package ggplot2--most modern and widely used.

**Lattice graphics** is built on top of the grid system, providing
high-level functions for all the common plot types. Plot output can be
saved as objects, edited and later plotted.

The three systems are, sadly, mostly incompatible, thus recommended to
choose one system for a project. ggplot2 is a good choice.

## Scatter plot

It is used to explore the relationship between  two continuous variables

### Using base R
Example: Does voters *Income* affect *Turnout* at the polls?

`plot(x)` to plot scatter plot. NA are omitted by default.

```{r}
# Deleting NA
obama_vs_mccain_turnout_non_na = obama_vs_mccain[!is.na(obama_vs_mccain$Turnout), ]
# plotting scatter plot
plot(obama_vs_mccain_turnout_non_na$Income, obama_vs_mccain_turnout_non_na$Turnout)
# or using with command
with(obama_vs_mccain_turnout_non_na, plot(Income, Turnout))
```

We can edit the plot by using various parameters as described in the help document `? plot` or `?? plot`

```{r}
with(obama_vs_mccain_turnout_non_na, plot(Income, Turnout, type = "p", main = "Scatter plot of Turnout versus Incoms",
                                          sub= "Obaba Vs McCain", xlab = "varage income", ylab = "Turnout percentage",
                                          col = "red", pch = 19, cex = 1.5 ))
```

Positive correlation between income and turnout

### Using ggplot2 package

Example: Does voters *Income* affect *Turnout* at the polls?

`ggplot()` to construct a plot. It takes a *dataframe* as its first argument and an *aesthetic* as its second. 

`geom_...()` to add e.g. points/line on the plot

```{r}
library(ggplot2)
gg1 = ggplot(obama_vs_mccain_turnout_non_na, aes(Income, Turnout, colour = Region)) + # colour points by region
geom_point(shape = 20)
print(gg1)
```

To draw individual plots for each region, we add a *facet* using the function `facet_wrap`.

```{r}
gg2 = gg1 + 
  facet_wrap(~ Region, ncol = 5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))   # to rotate x axis
plot(gg2)
  
```




## Line plot




## Histogram

## Bar chart



## Getting help
- [Posit Community, formally known as RStudio community](https://community.rstudio.com/)

- [Stack Overflow link for ggplot](https://stackoverflow.com/questions/tagged/ggplot2?sort=frequent&pageSize=50)

- [ggplot2 GitHub](https://github.com/tidyverse/ggplot2)

- [ggplot2 references for further reading](https://r4ds.hadley.nz/data-visualize)
