---
title: ' Summary statistics and visualisations'
author: "Bernard Silenou and Henrik Schanze"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
editor_options: 
  markdown: 
    wrap: 72
---

# Data Visualization

-   After importing and cleaning your data, the next aim before
    modelling is to understand your data.
-   Essential tools to attain this aim are *summary statistics* and
    *plots*.

## Goals

-   Be able to calculate a range of summary statistics on numeric data
-   Be able to draw plots and export out of R

# Summary statistics

Summary statistics are used to summarize data by precise numeric values
to develop an understanding of the data and communicate its key
information as simply as possible. The basic measurements often used for
this are

-   the arithmetic mean, to get an idea of the center of the data
    points,

-   the mean, to examine the data points in the middle,

-   the variance/standard deviation, to measure the range of data,

-   and the correlation coefficient, to examine the correlation of two
    variables in the data set.

To demonstrate these and other things we will use the `obama_vs_mccain`
data set, that is part of the `learningr` and contains the fractions of
people voting for Obama and McCain in the 2008 US presidential
elections. We want to get an idea of the votes for Obama.

```{r}
load(file ="data/obama_vs_mccain.rda")
votesObama = obama_vs_mccain$Obama
```

### Measure of location:

As a first measurement we calculate the arithmetic mean, which is the
sum of all data points divided by the number of data points, of the
votes by `mean()`.

```{r}
mean(votesObama)
```

As a second measure we calculate the median, i.e. the data point located
in the middle of all data points, by `mean()`.

```{r}
median(votesObama)
```

### Measure of range:

To get the smallest fraction of votes for Obama we could use `min()`.

```{r}
min(votesObama)
```

Similar to that we can get the highest amount of votes by `max()`.

```{r}
max(votesObama)
```

To get the data points that separates the set in different sections,
e.g. the first quartile being located above $25\%$ and below $75\%$ of
the other data points, we can use `quantile()` and use its second
argument to determine the fraction of data points below the value.

```{r}
quantile(votesObama, 0.25)
quantile(votesObama)
```

**Quick exercise**: Express `min()`, `max()` and `median()` via
`quantile()`.

```{r}
min(votesObama)
quantile(votesObama, )

max(votesObama)
quantile(votesObama, )

median(votesObama)
quantile(votesObama, )
```

### Measure of variation:

To get the range of our data, which is simply the smallest and largest
values of the data, we can use `range()`.

```{r}
range(votesObama)
```

For the standard deviation of your data, i.e. the average distance of
data points to its mean, we can use `sd()`.

`sd(x)`: standard deviation, square sd to get variance or use `var(x)`

```{r}
sd(votesObama)
```

Sometimes the variance, determined by `var()` is used instead of the
standard variation, which is just the square root of the variance.

```{r}
var(votesObama) 
sd(votesObama)**2
```

The interquartile range is the distance between the $25\%$ and $75\%$
percentile, and incorporates $50\%$ of the data including the median.
The calculated by one command, we can use `IQR()`.

```{r}
IQR(votesObama)
```

### Measure of position or count:

Sometimes we are interested in the data at certain positions. To access
the first value in the data we could use `[1]` following the data set.

```{r}
votesObama[1]
```

Since `length()` gives us the length of the data vector, we can use it
to access the last element.

```{r}
votesObama[length(votesObama)]
```

**Quick exercise**: what is the last last but one element in
`votesObama`?

```{r}

```

### Other summary statistics functions:

To get an statistical overview over the data, we can use `summary()`.

```{r}
summary(votesObama)
summary(obama_vs_mccain)
```

An alternative to `length()`, that is also useful, when operating with
data frames is `dim()`.

```{r}
length(votesObama)
dim(obama_vs_mccain)
dim(obama_vs_mccain)[1] # the same as nrow(obama_vs_mccain)
dim(obama_vs_mccain)[2] # the same as ncol(obama_vs_mccain)
```

If we have categorical data and want an overview about the distribution
among the categories we can use `table()`. For example we could examine
the amount of missing data by using `table()` after applying `is.na()`.

```{r}
table(is.na(votesObama))
table(is.na(votesObama), useNA = "always")
```

Get get a list of all answers given we could call `unique()` on the data
set. The length of its return also tells us how many different answers
there are.

```{r}
unique(obama_vs_mccain$Region)
length(unique(obama_vs_mccain$Region))
```

To calculate the correlation coefficient (pearson by default) of two
numeric vectors, we can use `cor()`, to get a hint if there might be a
link between both variables.

```{r}
cor(obama_vs_mccain$Obama, obama_vs_mccain$McCain)
```

Same wise we could calculate the covariance (also pearson by default) by
`cov()`.

```{r}
with(obama_vs_mccain, cov(Obama, McCain)) # the same as cov(obama_vs_mccain$Obama, obama_vs_mccain$McCain)
```

# Visualizations

## Systems of plots

R has three systems of graphics: *base*, *grid*, and *lattice*.

**Base graphics** is easy to get started with but hard to extend to new
graph types. It will be used for a greater part of this section.

**Grid graphics** provides greater flexibility but require writing much
lines of code. Example is package ggplot2--most modern and widely used.

**Lattice graphics** is built on top of the grid system, providing
high-level functions for all the common plot types. Plot output can be
saved as objects, edited and later plotted.

The three systems are, sadly, mostly incompatible, thus recommended to
choose one system for a project. ggplot2 is a good choice.

## Heat maps

When we want to visualize the connection of two categorical variables
our set of tools is quite limited. As an example we have a look at the
data set `mtcars` incorporating data about 32 cars. We want to visualize
the (possible) link between the number of cylinders `cyl` and number of
(forward) gears `gear`.

Since this data set is part of base **R**, we can use the R help to get
a better understanding what the columns mean.

```{r}
?mtcars
```

One of the most simple solutions is to use a cross table. To create one,
we can use `table()` and hand over the column names we are interested in
(you can hand in more than two, but the cases that this is useful are
quite limited). We could also hand over names for the row and column
names via `dnn`.

```{r}
table(mtcars$cyl, mtcars$gear, dnn = list("cylinder","gears"))
```

Depending on the number of categories and magnitude of numbers, spotting
an underlying correlation might be quite hard. A way to spot connections
faster is to color-code the cells of the cross table The resulting
graphic is called a heat map.

### Using base R

To transform a cross table into heat map we can give the cross table to
`heatmap()`.

```{r}
heatmap(table(mtcars$cyl, mtcars$gear))
```

Without any further arguments `heatmap()` also creates a dendrogram and
reorders the categories according to its clustering. To deactivate that
we have to set the arguments `Rowv` and `Colv` to `NA` and by setting
`revC=TRUE` we get the same cell order than in the cross table.

```{r}
heatmap(table(mtcars$cyl, mtcars$gear), Rowv = NA, Colv = NA, revC = TRUE,
        main = "Number of cylinders vs. number of gears",
        xlab = "Number of gears",
        ylab = "Number of cylinders")
```

From time to time it is reasonable to scale the color-coding according
to all values, column wise or row wise (the default). We could set this
by the argument `scale`.

```{r}
heatmap(table(mtcars$cyl, mtcars$gear), Rowv = NA, Colv = NA, revC = TRUE,
        scale = "row",
        main = "Number of cylinders vs. number of gears",
        xlab = "Number of gears",
        ylab = "Number of cylinders")
```

### Using ggplot2 package

First of all we have to load the `ggplot2` package. If `ggplot2` is not
installed, please use `install.packages("ggplot2")` to do so.

```{r}
library(ggplot2)
```

Creating a heat map with `ggplot2` requires a little bit extra work, but
it enables us to adjust the graphic in more detail. We also start with
the initial cross table, but turn it into a data frame

```{r}
df_cars <- data.frame(table(mtcars$cyl, mtcars$gear, dnn = list("cylinder","gears")))

df_cars
```

Now we can hand over that data frame to `ggplot()` and specify in
`aes()` which variable (from the data frame) should be on the `x` axis,
on the `y` axis and which numbers should the cells be filled with
(`fill`). Finally we can add title and labels as before.

```{r}
ggplot(df_cars, aes(x = gears, y = cylinder, fill = Freq)) +
  geom_tile() +
  labs(main = "Number of cylinders vs. number of gears",
       x = "Number of gears",
       y = "Number of cylinders")
```

Adjusting the scaling of this heat map is also a little bit more complex
than before, since we have to apply `scale()` to the initial table. The
argument `center` determines the new central value and the argument
`scale` determines the way we scale the values. `scale = FALSE` does no
scaling, `scale = TRUE` gives row scaled values and `scale = c(15,12,5)`
results (in this case) in a column scaling.

```{r}
df_cars_scaled <- data.frame(scale(
  table(mtcars$cyl, mtcars$gear, dnn = list("cylinder","gears")),
  center = FALSE, scale = c(15,12,5)))

ggplot(df_cars_scaled, aes(x = gears, y = cylinder, fill = Freq)) +
  geom_tile() +
  labs(main = "Number of cylinders vs. number of gears",
       x = "Number of gears",
       y = "Number of cylinders")
```

## Bar chart

A bar chart displays the amount of of a discrete categorical data via
the height or length of a bar. Using a bar chart is therefore a visual
alternative for using a table.

As an example we will again use the data set `mtcars` which includes
information about 32 different cars, e.g. the number of cylinders `cyl`.

```{r}
mtcars
```

```{r}
table(mtcars$cyl, dnn = list("cylinders"))
```

### Using base R

As we can see the data set `mtcars` lists different cars in their rows
including information about the number of cylinders `cyl`. To create a
bar chart depicting the number of cars with a certain amount of
cylinders, we can not use this raw data, since this information is not
directly part of the data set.

To create such a bar chart we first have to extract the required
information out of the data set. Luckily `table()` does exactly what we
want. If we use it and turn the result via `data.frame()` into a data
frame (so we have access to the row listing the counts), we can use
`barplot()` to create the respective bar chart.

```{r}
cyl_table <- data.frame(table(mtcars$cyl, dnn = list("cylinders")))
cyl_table
barplot(height = cyl_table$Freq, names = cyl_table$cylinders,
        main = "Amount of cars with respective number of cylinders",
        xlab = "Number of cylinders")
```

### Using ggplot2 package

If not already done, please load the `ggplot2` package.

```{r}
library(ggplot2)
```

Using `ggplot2` has the advantage that we can feed our data directly to
the 'plotting'-function without count the values of which amount we are
interested in 'manually'. We can give our data set to `ggplot()`, and
specify in `aes()` that we are interested to plot `cyl` a long the
x-axis, by setting `x = cyl`. Since the numbers in `cyl` represent
categories and not (directly) numbers, we have to use `as.factor()` on
the variable `cyl`. After this initial step, we could create the bar
chart with adding `geom_bar()`, where we can specify details about the
bar chart, and `labs()` to set all kinds of labels for our graphic.

```{r}
ggplot(mtcars, aes(x = as.factor(cyl))) +
  geom_bar() +
  labs(title = "Amount of cars with respective number of cylinders",
       x = "Number of cylinders",
       y = "Number of cars")
```

## Histogram

Histograms are a very basic way to display an approximation of the
distribution of a given variable. In this example we want visualize the
data in `airquality` that includes measurements in New York from May to
September 1973.

First we have a short look into the data.

```{r}
head(airquality)
```

In the following we want to visualize the distributions of wind speed
(measured in miles per hour). To get a first impression of the range of
that variable we use `summary()`.

```{r}
summary(airquality$Wind)
```

### Using base R

We want to use histograms to get a visual impression of the recorded
wind speeds. A histogram uses bins to collect ranges of data and depicts
the number of data points in each bin by its height. In base **R** we
can easily create such a plot by using `hist()`. Without additional
arguments **R** will distribute the 'bins' on its own, but we can also
manually adjust the number or width of the used 'bins' by the argument
`breaks`.

```{r}
hist(airquality$Wind,
     xlab = "Wind speed")

hist(airquality$Wind, breaks = 25,#c(0,seq(5,15,2), 21),
     xlab = "Wind speed")
```

Since it would be interesting to compare the distribution of wind speeds
for different months, we want to directly compare two histograms with
each other.

```{r}
hist(airquality[airquality$Month == 5,]$Wind,
     breaks = c(0,seq(5,15,2), 21), col = rgb(1,0,0,0.5),
     xlab = "Wind speed", main = "")

hist(airquality[airquality$Month == 8,]$Wind,
     breaks = c(0,seq(5,15,2), 21), col = rgb(0,0,1,0.5),
     add = TRUE)
```

### Using ggplot2 package

We want to show the way histograms could be created with `ggplot2`. In
case you have not loaded `ggplot2` please execute the following line.

```{r}
library(ggplot2)
```

We first have to tell `ggplot2` what data we are using by `ggplot()` and
specifying that we want to display `Wind` on the x-axis. Afterwards we
can create the histogram with `geom_histogram()`. As before not giving
any information about the wanted bins leads to `ggplot` deciding how to
distribute the bins according to the data. We can either choose the
number of bins by `bins`, the width of the bins by `binwidth` or specify
where to separate between bins with `breaks`. Each of these arguments
overwrites the use of the former.

```{r}
ggplot(airquality, aes(x = Wind)) +
  geom_histogram()

ggplot(airquality, aes(x = Wind)) +
  geom_histogram(breaks = c(0,seq(5,15,2), 21))
```

We can also use `ggplot2` compare two (or more) histograms with each
other. In contrast to base R we do this in one single command using
`Month` as a factor (notice that we have selected the data for only two
different months in the first argument).

```{r}
ggplot(airquality[airquality$Month %in% c(5,8),],
       aes(x = Wind, fill = as.factor(Month))) +
  geom_histogram(position="identity", alpha=0.5, breaks=c(0,seq(5,15,2), 21)) +
  scale_fill_manual(values = c(rgb(1,0,0,0.5), rgb(0,0,1,0.5)))
```

## Box plot

A box plot is a visualization of the distribution of numerical data.
Will we use the data set `mtcars` which incorporates fuel consumption
and ten aspects design and performance for 32 different cars. First we
have a short look into our data.

```{r}
head(mtcars)
```

As we have seen before we can use `summary()` to get a (numerical)
overview about the distribution of the fuel consumption `mpg` (miles per
gallon).

```{r}
summary(mtcars$mpg)
```

That may answer some of our question regarding the over all fuel
consumption of the cars in the data set, but it es not very easy to gasp
and gets very overwhelming if we want to compare different distributions
of fuel consumption (e.g. distinguished by number of cylinders).

### Using base R

Using a box plot for visualizing puts all the above information in one
graphic.

```{r}
boxplot(mtcars$mpg)
```

Since the single plot includes all the information from before it needs
some knowledge to be understood completely.

-   The Median is marked by the thick black line inside the grey box.

-   The 1st and 3rd quartile are the lower and upper end of the grey
    box, therefore $50\%$ of the values, including the median, are
    located inside the grey box.

-   The two antennas/whiskers give the range including (roughly) $95\%$
    of the data points.

-   Values outside this range are sometimes depicted as singular data
    points.

We can also use box plots to quickly compare the distribution of the
fuel consumption for different number of cylinders `cyl` or number of
gears `gear`.

To do that we have to give `boxplot` a so called formula describing the
relation we are interested in. A formula connects the (numeric) variable
we are interested in (e.g. `mgp`) via `~` with the grouping variables,
such as `cyl` or `gear`, usually as factors.

```{r}
boxplot(mpg ~ cyl, data = mtcars)
boxplot(mpg ~ gear, data = mtcars)
```

As before with other plots we have some basic options to manipulate the
appearance of the plots, e.g. add labels.

```{r}
boxplot(mtcars$mpg,
        main="Boxplot von mpg gruppiert nach cyl",
        xlab="Zylinder (cyl)", ylab="Miles per Gallon (mpg)")
```

We can also manually adjust the names of the categories.

```{r}
boxplot(mpg ~ vs, data = mtcars, names  = c("V-shaped", "Straight"))
```

### Using ggplot2 package

Again for further and more specific adjustment we could use `ggplot2`
instead of base **R**.

```{r}
library(ggplot2)
```

As with nearly every plot done with `ggplot` we start with making
`ggplot` familiar with the data we are using and which parts we want to
visualize. As before we want to use the data set `mtcars` and display
the fuel consumption `mpg`. Since we want to display the box plot
vertically we assign the variable `mpg` to the y-axis.

```{r}
ggplot(mtcars, aes(y = mpg)) + 
  geom_boxplot() + 
  labs(title = "Boxplot fuel consumption",
       y = "Miles per Gallon (mpg)")
```

If we want to plot the data grouped in the different numbers of
cylinders `cyl` we have to make sure that the variable we use to group
the data is a factor. We can achieve this by using `as.factor()`.

```{r}
ggplot(mtcars, aes(x = as.factor(cyl), y = mpg)) +
  geom_boxplot() +
  labs(title = "Boxplot fuel consumption",
       x = "Number of cylinder (cyl)",
       y = "Miles per Gallon (mpg)")
```

## Line plot

For continuous data depending on time we often use a line plot, since
the the time points often are not equally distributed and a linear
interpolation between data points is a reasonable guess. For
demonstration we will use the data set `crab_tag` from the `learningr`
package.

```{r}
# Extracting data from package
#data(crab_tag, package = "learningr")
load(file = "data/crab_tag.rda")
crab_tag_daylog = crab_tag$daylog 
str(crab_tag_daylog)
```

### Using base R

We want to visualize `Max.Depth` depending on `Date` via a line plot. To
do this in base R we can simply use `plot()` with the first variable as
the first argument `x` and the second variable as the second argument
`y`. To get a line plot we also have to set `type = "l"`, since `plot()`
is capable of many different kinds of plots.

```{r}
# Drawing line plot
plot(crab_tag_daylog$Date, crab_tag_daylog$Max.Depth, type = "l")
```

Since the `Max.Depth` describes a depth it would be nice to visualize
this accordingly. To adjust the plot, we can simply assign a `-` in
front of the data and to determine the scaling of the y-axis, we can use
`ylim`.

```{r}
plot(crab_tag_daylog$Date, -crab_tag_daylog$Max.Depth, 
     type = "l", ylim = c(-max(crab_tag_daylog$Max.Depth)-10, 10)) 
```

We can draw multiple lines in the same plot, e.g. if we want to
visualize `Max.Depth` and `Min.Depth` simultaneously. We just create the
line plot for the first variable and and create the second line in the
plot with `line()`. To distinguish the lines we could change their
colors by the argument `col`.

```{r}
plot(crab_tag_daylog$Date, -crab_tag_daylog$Max.Depth, 
     type = "l", ylim = c(-max(crab_tag_daylog$Max.Depth)-10, 10))
# Adding the lines for min depth on the existing plot of max depth

lines(crab_tag_daylog$Date, -crab_tag_daylog$Min.Depth, col = "blue")
```

### Using ggplot2 package

We can do the same with `ggplot2`. If not done yet, load the package
with the following line.

```{r}
library(ggplot2)
```

We first hand `ggplot()` the data we want to use and tell it in `aes()`
that we want `Date` along the x-axis. In the second step we add the line
via `geom_line()` and the variable it should show. If we want to add
another line depicting another variable we simply can add the same
command with a different variable. As before we can give different
colors to the lines with the argument `colour`.

```{r}
ggplot(crab_tag_daylog, aes(x = Date)) +
geom_line(aes(y = -Max.Depth), colour = "#a83b48") + # google "colour picker" to get the codes
geom_line(aes(y = -Min.Depth), colour = "darkgrey")
```

## Scatter plot

If we want to visualize the relationship of two continuous variables a
practicable thing to do is to use a scatter plot. Each variable is
assigned the role of a dimension and the resulting data points are then
depicted in the resulting two-dimensional space (of course we could do
the same with more than two variables, but using more then three becomes
harder to visualize properly).

To demonstrate this plot and its creation we want to come back to the
data set `obama_vs_mccain` from the begining of the lecture.

```{r}
load(file ="data/obama_vs_mccain.rda")
```

### Using base R

If we are interested in the correlation of `Income` and `Turnout`, so to
speak the question if the income influences the turnout of the poll, we
can use a scatter plot to get a visual impression, whether this is
plausible or not.

In base R we can simply use `plot()` to create a scatter plot. Since it
has no build in mechanism to deal with missing values, we first have to
clean our data of these. Afterwards we can hand over the data we want to
use for the x-axis `x` and y.axis `y` respectively.

```{r}
# Deleting NA
obama_vs_mccain_turnout_non_na = obama_vs_mccain[!is.na(obama_vs_mccain$Turnout), ]
# plotting scatter plot
plot(obama_vs_mccain_turnout_non_na$Income, obama_vs_mccain_turnout_non_na$Turnout)
# or using with command
with(obama_vs_mccain_turnout_non_na, plot(Income, Turnout))
```

We can edit the plot by using various parameters as described in the
help document `?plot` or `??plot`. For example we can change the color
via `col`.

```{r}
plot(obama_vs_mccain_turnout_non_na$Income,
     obama_vs_mccain_turnout_non_na$Turnout,
     col = "red", pch = 20, cex = 3.5,
     main = "Scatter plot of Turnout versus Incoms",
     sub= "Obaba Vs McCain", xlab = "varage income", ylab = "Turnout percentage")
```

### Using ggplot2 package

Alternatively to base R we can use `ggplot2` to create a scatter plot.
If not already done, please load the `ggplot2` package by the following
line.

```{r}
library(ggplot2)
```

To create a scatter plot showing the relation of `Income` and `Turnout`
we have to first hand the data set of `ggplot()` and give details about
the variables we want to show in `aes()`. Followed by that we use
`geom_point()` to create the points of the scatter plot. We can change
details of the plot by arguments of the function, e.g. the size of the
dots by `size`.

```{r}
ggplot(obama_vs_mccain_turnout_non_na, aes(x = Income, y = Turnout)) +
geom_point(size = 5)
```

We can use coloring to easily depict a third variable in our two
dimensional plot, e.g. to examine the influence of `Region` on the other
variables.

```{r}
ggplot(obama_vs_mccain_turnout_non_na, aes(x = Income, y = Turnout, color = Region)) +
geom_point(size = 3)
```

To draw individual plots for category in `Region`, we add a *facet*
using `facet_wrap()`.

```{r}
ggplot(obama_vs_mccain_turnout_non_na, aes(x = Income, y = Turnout)) +
  geom_point(size = 3) + 
  facet_wrap(~ Region, ncol = 5) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))   # to rotate x axis
```

## Additional visualizations using ggplot2 package

-   [ggplot2 references for further
    reading](https://r4ds.hadley.nz/data-visualize)
-   [ggplot2 references
    manual](https://cloud.r-project.org/web/packages/ggplot2/ggplot2.pdf)

# Getting help

-   [Posit Community, formally known as RStudio
    community](https://community.rstudio.com/)

-   [Stack Overflow link for
    ggplot](https://stackoverflow.com/questions/tagged/ggplot2?sort=frequent&pageSize=50)

-   [ggplot2 GitHub](https://github.com/tidyverse/ggplot2)

-   [ggplot2 references for further
    reading](https://r4ds.hadley.nz/data-visualize)
