---
title: "Homework — Introduction to R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

- Open this file in **RStudio**.
- You do **not** need to know anything about R Markdown.
- Run code chunks by clicking the green **► Run** button.
- Write your answers directly in the code chunks.

---

# 1. Calculator and simple functions

### Task 1
Run the following calculations.  
Then create **your own** calculation.

```{r}
345 + 5
60/5*10
abs(-12)
```

# Your own calculation:

```{r}
3*4/8+5
2^2
```


# 2. Creating objects

### Task 2

Create an object x containing the number 42.
Print it.
Then add 10 to it and store it in an object y.

```{r}
x <- 42
y <- x + 10
```


### Task 3

Create a vector y with the following values:
-3, -1, 0, 2, 4

Then:

Print y

Apply the function abs() to y

Calculate the mean of y


```{r}
y <- c(-3, -1, 0, 2, 4)
print(y)
y
abs(y)
mean(y)

```



# 3. Inspecting data frames

We will use the built-in dataset mtcars.


### Task 4

Load mtcars and inspect its structure.

```{r}
# Load:
data(mtcars)
# Inspect:
str(mtcars)

```


### Task 5

Extract from mtcars:

1. The second column

2. The fifth row

3. The value in row 3, column 4

4. The column hp using $

```{r}

mtcars[,2]
mtcars[5,]
mtcars[3,4]
mtcars$hp

```


# 4. Simple data wrangling with tidyverse

We use the dataset iris.

```{r}
library(tidyverse)
data(iris)
```

### Task 6

Use filter() to extract all plants with sepal length (Sepal.Length) below / equal to 4.9.

```{r}
iris %>% filter(Sepal.Length <= 4.9)

```


### Task 7

Extract all plants with a Petal Width of 0.2 AND belonging to Species Setosa. 
```{r}
iris %>% filter(Petal.Width == 0.2 & Species == "setosa")

```


### Task 8

By using the pipe (%>%),
- first create variables "Sepal.Length.mm", "Sepal.Width.mm", "Petal.Length.mm", "Petal.Width.mm";
- by converting the values in the original variables that are in cm to mm;
- afterwards select Species Setosa;
- save the results in a new object with a reasonable name.
 
```{r}
iris_mm_setosa <- iris %>% 
  mutate(Sepal.Length.mm = Sepal.Length*10,
         Sepal.Width.mm = Sepal.Width*10,
         Petal.Length.mm = Petal.Length*10,
         Petal.Width.mm = Petal.Width*10) %>%
  filter(Species == "setosa")

```




### Task 9

Select the rows Petal.Length and Petal.Width from your last dataset by using a tidyverse function.
To find the right function, use search machines or AI.
Back in R, open the help page of the function with ?.

```{r}
?select

iris_mm_setosa %>% select(Petal.Length, Petal.Width)

```



